Update of the ozone dry deposition in the OsloCTM3 v1.0

### 2018-10-09 DryDep.tex
Finished revision of first draft of introduction. May be tagged.

\begin{abstract}
  Since the industrial revolution, tropospheric background ozone concentrations have been increasing strongly mainly in the northern hemisphere. In recent years, the number of episodes of peak concentrations especially in summertime has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in particular may even affect food security in the future. Most chemistry-transport models (CTMs) and chemistry-climate models (CCMs) tend to overestimate surface ozone values. This may be partly due to uncertainties in the distribution and amount of precursor species and stratosphere-troposphere exchange as well as scavenging processes removing ozone from the atmosphere. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. Dry deposition velocities are typically calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these conceptual dry deposition velocities are hard to measure explicitly. In this paper, we present an update on the general dry deposition scheme implemented in the OsloCTM3v1.0. We evaluate the resulting modelled ozone dry deposition with respect to obervations and multi-model studies. We will also estimate the impact of a set of parameters on the modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition decreases with respect to the previous version, leading to an increase in surface ozone of up to ??\,\unit{\%}. {\bf TODO: Finsish the results section as soon as there are results.}
\end{abstract}

@@

In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about 3--15\,\unit{\%} \citep{PJ:Ainsworth2017} as well as 7\,\unit{\%} loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants in particular may even affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\

@@

Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolysed. The resulting atomic oxygen reacts then with \chem{O_2} (and any other available co-reactant $M$) to form an ozone molecule

@@

Similar reaction mechanisms involving volatile organic compounds (VOCs) exist \citep{ACP:Monks2015}. Another source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Based on observations, STE might only amount to roughly 10\,\unit{\%} ($550 \pm 140$\,\unit{Tg a^{-1}}) of the total global ozone budget in the troposphere, while ozone from chemical production is estimated to be 5000\,\unit{Tg a^{-1}} \citep{ACP:Monks2015}. Ozone is removed from the atmosphere by photochemical reactions or scavenging processes. Major sinks are photolysis followed by a reaction with water vapor to from \chem{OH}

@@

or de-nitration reaction via

@@

  \chem{NO_x} + \chem{O_3} \rightarrow \chem{NO_{x+1}} + \chem{O_2}.

@@

A minor sink which is related to dry deposition are so called ozone depleting events. They occur in spring-time in the polar boundary layer where an outburst of bromine monoxide \chem{BrO} (so called bromine explosion) leads to a rapid depletion of surface ozone \citep{JGR:Oltmans1981,GRL:Bottenheim1986,Nat:Barrie1988,JGR:Bottenheim2006}. Various schemes ranging from bulk-snow parameterizations \citep{ACP:Toyota2011,GMD:Falk2018} to detailed in-snow \citep{ACP:Toyota2014a}, and aerosol chemistry \citep{ACP:Yang2010} have been successfully applied to different types of atmospheric models. Although these events are important to understand surface ozone abundance in Arctic regions, we have not implemented any parameterization of these processes in the OsloCTM3.\\
Since ozone is highly reactive, its global mean life-time in the troposphere is roughly 22 days but ranges between a few days in the tropical boundary layer to up to one year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day (maximum $\sim$15:00 local time), season (mid-June maximum), altitude, location \citep{ACP:Schnell2015}, or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from 10\,\unit{ppb} over the tropical Pacific to 100\,\unit{ppb} in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observation as well as validation and intercomparison of models. At the observational side, there is only a limited number of long-term ozone observations, mainly restricted to European sites. Amongst these, only three display a statistical significant trend, indicating a doubling of tropospheric ozone since the onset of the industrialization \citep[Chapter 2]{IPCC2013}. But especially the very low pre-industrial ozone abundance cannot be reproduced by the likes of most models, while there is a general tendency to underestimate tropospheric ozone among the participating models in the Atmospheric Chemistry and Climate Model Intercomparison Project (ACCMIP) \citep[Chapter 8]{IPCC2013}.\\

@@

wherein aerodynamic $R_a$, quasi laminar $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gaseous specie to specie and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years, e.g., dry deposition of ozone to ocean \citep{ACP:Luhar2017} and stomatal conductance as part of canopy conductance \citep{PP:Buckley2017} or updates on the parameters have been provided from model validation with observation \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. The variaty of differing formulations leads to a wide spread in model intercomparisons \citet{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition.\\

In Section~\ref{sec:model_des}, we will briefly describe the OsloCTM3, give a detailed account of the updates of the dry deposition scheme (Section~\ref{subsec:DryDep}) as well as present the pre-processing of meteorological input data to compute necessary input to the dry deposition scheme such as begin and duration of greening season and photosynthetic photon flux density (Section~\ref{subsec:pre-pro}). In Section~\ref{sec:eval}, we present sensitivity tests with respect to a manifold of parameters in the drydeposition scheme and validate our results with results from the modelintercomparison of \citet{ACP:Hardacre2015} and to observations.
{\bf TODO: Outlook on paper.}

@@

historical anthropogenic and biomass burning emissions \citep{Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

\subsection{Comparison with recent multi-model evaluation}
\label{subsec:model}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}.

@@

\section{Discussion}

@@

\conclusions  %% \conclusions[modified heading if necessary]
What did we gain?

>>>>

\begin{abstract}
  Since the industrial revolution, tropospheric background ozone concentrations have been increasing in the northern hemisphere. In recent years, the number of episodes of peak concentrations has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in combination with advancing climate change may affect food security in the future. While the future scenarios in themselves are uncertain, there are limiting factors constraining the accuracy of surface ozone modeling also at present: The distribution and amount of precursor/decomposing species, the stratosphere-troposphere exchange as well as scavenging processes. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. The conceptual dry deposition velocities are calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these are hard to measure explicitly. In this paper, we present the update from an empirical dry deposition parameterization to a more process based implementation in the OsloCTM3. Our focus lies mainly on a stomatal conductance-based description of the canopy resistance. We evaluate the resulting modeled ozone dry deposition with respect to observations and multi-model studies. We will also estimate the impact of a set of parameters on the modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition decreases with respect to the previous version, leading to an increase in surface ozone of up to ??\,\unit{\%}. {\bf TODO: Finish the results section as soon as there are results.}
\end{abstract}

@@

In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about 3--15\,\unit{\%} \citep{PJ:Ainsworth2017} as well as 7\,\unit{\%} loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants may affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\

@@

Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolyzed. The resulting atomic oxygen reacts then with \chem{O_2} (and any other available co-reactant $M$) to form an ozone molecule

@@

Similar cycles involving volatile organic compounds (VOCs) exist \citep{ACP:Monks2015}. Another source of tropospheric ozone is downward transport from the stratosphere via stratosphere-troposphere exchange (STE) \citep{WMO2014}. Based on observations, STE might only amount to roughly 10\,\unit{\%} ($550 \pm 140$\,\unit{Tg a^{-1}}) of the total global ozone budget in the troposphere, while ozone from chemical production is estimated to be 5000\,\unit{Tg a^{-1}} \citep{ACP:Monks2015}. Ozone is removed from the atmosphere by photochemical reactions or scavenging processes. Major sinks are photolysis followed by a reaction with water vapor to from \chem{OH}

@@

de-nitration reactions via

@@

  \chem{NO_x} + \chem{O_3} \rightarrow \chem{NO_{x+1}} + \chem{O_2},

@@

and dry deposition. We will come back to the latter later in this section and cover the implemented scheme in more detail in Section~\ref{subsec:DryDep}. A dry deposition related sink limited to Arctic regions are so called ozone depleting events. They occur in spring-time in the polar boundary layer where an outburst of bromine monoxide \chem{BrO} (so called bromine explosion) leads to a rapid depletion of surface ozone \citep{JGR:Oltmans1981,GRL:Bottenheim1986,Nat:Barrie1988,JGR:Bottenheim2006}. Various schemes ranging from bulk-snow parameterization \citep{ACP:Toyota2011,GMD:Falk2018} to detailed in-snow \citep{ACP:Toyota2014a}, and aerosol chemistry \citep{ACP:Yang2010} have been successfully applied to different types of atmospheric models but do not yet cover the full range of observed events. Although these ozone depleting events are important to understand surface ozone abundance in Arctic regions, we have not implemented any parameterization of these processes in the OsloCTM3 as of now.\\
Since ozone is highly reactive, its global mean life-time in the troposphere is roughly 22 days but ranges between a few days in the tropical boundary layer to up to one year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day (maximum $\sim$15:00 local time), season (mid-June maximum), altitude, location \citep{ACP:Schnell2015}, or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from 10\,\unit{ppb} over the tropical Pacific to 100\,\unit{ppb} in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observation as well as validation and intercomparison of models. At the observational side, there is only a limited number of long-term ozone observations, mainly restricted to European sites. Amongst these, only three display a statistical significant trend, indicating a doubling of tropospheric ozone since the onset of the industrialization \citep[Chapter 2]{IPCC2013}. But especially the very low pre-industrial ozone abundance cannot be reproduced by the likes of most models. Among the participating models in the Atmospheric Chemistry and Climate Model Intercomparison Project (ACCMIP), there is a general tendency to underestimate tropospheric ozone burden (e.g., $10-20$\,\unit{\%} negative bias at 250\,\unit{hPa} in the southern hemisphere (SH) tropical region) \citep[Chapter 8]{IPCC2013}. With respect to surface ozone, \citet{ACP:Schnell2015} conclude that all ACCMIP models, which reported hourly surface ozone, tend to overestimate surface ozone values in North America and Europe in comparison with available observations. A key to fathom these slightly contradicting results may lie in the used dry deposition schemes.\\

@@

wherein aerodynamic $R_a$, quasi laminar $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gas to gas and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years. \citet{ACP:Luhar2017} have validated ozone dry deposition to the ocean with respect to three different formulations of surface resistances. An update on the ozone surface resistance over snow and ice covered surfaces has been provided from combined model and observation studies \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. Canopy conductance is formulated at the single-leave level (stomatal conductance) for various plant function types (PFT) as well as single species based on empirical studies \citep{PTRS:Jarvis1976, BallBerry1987, ACP:Simpson2012, ICP:MappingManual2017}. But progress has also been made on process oriented modeling of stomatal conductance \citep{PP:Buckley2017}. The variety of differing formulations and choices of parameters leads to a wide spread in model intercomparisons \citep{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition \citep{ACP:Monks2015}.\\

In Section~\ref{sec:model_des}, we will briefly describe the OsloCTM3, give a detailed account of the new dry deposition scheme (Section~\ref{subsec:DryDep}) as well as present pre-processing of meteorological input data to compute necessary input to the dry deposition scheme such as begin and duration of greening season (GDAY, GLEN) and photosynthetic photon flux density (PPFD) (Section~\ref{subsec:pre-pro}). In Section~\ref{sec:eval}, we present sensitivity tests with respect to a manifold of parameters in the dry deposition scheme (Section~\ref{subsec:sens}) and validate our results with results from the multi-model intercomparison of \citet{ACP:Hardacre2015} (Section~\ref{subsec:model}) and to observations both at the surface as well as in the free troposphere (Section~\ref{subsec:obs}). In Section~\ref{sec:disc}, we will summarize our results and draw conclusions for further development of the model (Section~\ref{sec:conc}).
{\bf TODO: Precise outlook on paper when writing is done.}

@@

historical anthropogenic and biomass burning emissions \citep{ACP:Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

\subsection{Comparison with recent multi-model comparison}
\label{subsec:model}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}.

@@

\section{Discussion}
\label{sec:disc}

@@

\conclusions  %% \conclusions[modified heading if necessary]
\label{sec:conc}
What did we gain?
